Statistical Reporting Inconsistencies in Experimental Linguistics

### 2025-07-19 manuscript/manuscript.qmd
prepared script for manual assessment

```{=tex}
\begin{equation*}
  \tikzmarknode{node1}{(1) }
  \eqnmarkbox[gray]{node2}{F}
  \eqnmarkbox[orange]{node3}{(1,66)}
  \tikzmarknode{node4}{=}
  \eqnmarkbox[purple]{node5}{3.88}
  \tikzmarknode{node6}{, }
  \eqnmarkbox[blue]{node7}{p < .05}
\end{equation*}

\annotate[yshift=1em]{left}{node2}{name of test} 
\annotate[yshift=-1em]{below,left}{node3}{degrees of freedom} 
\annotate[yshift=1em]{right}{node5}{test statistic} 
\annotate[yshift=-1em]{below,right}{node7}{p-value}
```

@@

@@

## Analysis and research questions
The aims of this study were explicitly exploratory and hypothesis generating, thus analyses remain merely descriptive by reporting on the proportion of articles/tests that are statistically (in)consistent.
Our investigation was set out to explore the following research questions:
(1) How prevalent are statistical inconsistencies (and decision inconsistencies) in our sample? (see @sec-prevalent)
(2) Do inconsistency rates vary across journals and/or their publication year? (see @sec-stable)
And (3) is there evidence for bias, i.e. are processes that result in inconsistencies more likely to produce lower p-values? (see @sec-bias)

@@

@@

Journal editors could explicitly recommend consistency checks with algorithms such as Statcheck during peer review, a practice that has been taken up on by several journals from neighboring disciplines (Psychological Science[^1], Advances in Methods and Practices in Psychological Science[^2], Stress & Health @barber2017meticulous). 
Editors could also demand, recommend or at least encourage data sharing for publication in their journal. 
Data sharing policies have been shown to substantially increase the reproducibility of analyses [e.g., @hardwicke2018data; @laurinavichyute2022share] and a number of linguistic journals have already implemented such policies, including journals within our sample. 
Having said that, open data policies which for example the Journal of Memory and Language introduced in 2018 did not seem to have affected the proportion of statistical inconsistencies after their introduction. 
The inconsistency rates for JML (and other journals) were rather stable across time, so open data alone might not resolve the issue without further changes to the research eco-system.

>>>>

\begin{equation*}
  \tikzmarknode{node1}{(1) }
  \eqnmarkbox[gray]{node2}{F}
  \eqnmarkbox[orange]{node3}{(1,66)}
  \tikzmarknode{node4}{=}
  \eqnmarkbox[purple]{node5}{3.88}
  \tikzmarknode{node6}{, }
  \eqnmarkbox[blue]{node7}{p < .05}
\end{equation*} \annotate[yshift=1em]{left}{node2}{name of test} \annotate[yshift=-1em]{below,left}{node3}{degrees of freedom} \annotate[yshift=1em]{right}{node5}{test statistic} \annotate[yshift=-1em]{below,right}{node7}{p-value}

@@

suppressWarnings(suppressMessages(suppressPackageStartupMessages({
source("../scripts/02_sampling_for_manual_inspection.R")
})))

@@

## Analysis and research questions

The aims of this study were explicitly exploratory and hypothesis generating, thus analyses remain merely descriptive by reporting on the proportion of articles/tests that are statistically (in)consistent.
Our investigation was set out to explore the following research questions: (1) How prevalent are statistical inconsistencies (and decision inconsistencies) in our sample?
(see @sec-prevalent) (2) Do inconsistency rates vary across journals and/or their publication year?
(see @sec-stable) (3) is there evidence for bias, i.e. are processes that result in inconsistencies more likely to produce lower p-values?
(see @sec-bias) And (4) do decision inconsistencies matter, i.e. are they merely typos or do they lead to erroneous theoretical interpretations?
(see @sec-matter)

@@

## Do decision inconsistencies matter? {#sec-matter}

Now the question arises, do the inconsistencies theoretically matter.
This is a difficult question to empirically address because experimental behavioral research is often characterized by weak links between theory and predictions [@scheel2021hypothesis], making it difficult to even extract hypotheses and whether hypotheses have been corroborated or not from published articles [@scheel2022most].
To gain some insight, we approximate an answer to this questions as follows: First, we extracted a random subsample (15%, n = `{r} nrow(xdata_sample)`) of articles which contain at least one decision inconsistencies, equally distributed across publication years.
We restricted the subsample to only those decision inconsistencies that involve either a larger (\>) or small (\<) comparison of the p-value to allow unambiguous interpretation of the results.
We then manually extracted the relevant sentence which the test statistic is embedded in and assessed whether the authors interpreted the reported p-value correctly (a true decision inconsistency) or incorrectly (the decision inconsistency is likely a typo).
This ratio will let us approximate the true decision inconsistency rate.
We consider true decision inconsistencies theoretically relevant.
E.g. if an author claims that two groups are different (or not) based on an erroneously reported significant test, we consider this claim to potentially have theoretical consequences.
Out of the `{r} nrow(xdata_sample)` assessed cases, we coded ## (##%) as genuine typos, e.g. the authors interpreted the results in line with the computed p-value and not the erroneously reported p-value.
We coded ## (##%) cases as unclear, i.e. even after assessing the manuscript in detail, we could not interpret how the p-value was interpreted. 
This leaves ## (##%) of tests being coded as true decision inconsistencies that potentially have theoretical consequences.

@@

Journal editors could explicitly recommend consistency checks with algorithms such as Statcheck during peer review, a practice that has been taken up on by several journals from neighboring disciplines (Psychological Science[^1], Advances in Methods and Practices in Psychological Science[^2], Stress & Health @barber2017meticulous). Editors could also demand, recommend or at least encourage data sharing for publication in their journal
. Data sharing policies have been shown to substantially increase the reproducibility of analyses [e.g., @hardwicke2018data; @laurinavichyute2022share] and a number of linguistic journals have already implemented such policies, including journals within our sample
. Having said that, open data policies which for example the Journal of Memory and Language introduced in 2018 did not seem to have affected the proportion of statistical inconsistencies after their introduction
. The inconsistency rates for JML (and other journals) were rather stable across time, so open data alone might not resolve the issue without further changes to the research eco-system
.
